---
title: "Reproducible Research: Peer Assessment 1"
author: "Paolo Coraggio"
date: "07/01/2020"
output: 
    html_document:
        keep_md: true
---

```{r setup, include=FALSE}
library(lubridate)
library(dplyr)
library(ggplot2)

knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This report analyse data taken from a personal activity monitoring device. The device collects data at 5 minutes intervals through out the day.

The report is organised in the following sections:

-   [Loading and preprocessing the data](#loading-and-preprocessing-the-data)
-   [Plotting the histogram of total number of steps taken each day and calculating mean and median](#what-is-mean-total-number-of-steps-taken-per-day)
-   [Time series of the average number of step taken and calculating the time interval with maximum number of steps](#what-is-the-average-daily-activity-pattern)
-   [A strategy for imputing missing data and plotting a histogram of total number of steps taken each day with the new imputed data](#imputing-missing-values)
-   [A comparison of the average number of steps taken in the different intervals between weekdays and weekends](#are-there-differences-in-activity-patterns-between-weekdays-and-weekends)

## Loading and preprocessing the data

The data can be downloaded from their online repository or, if available, from a local directory. It also creates a `figure` subfolder to store the produced images. 

```{r, echo=TRUE}
if(!file.exists("./data/activity.csv"))
{
    link <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
    temp <- tempfile()
    download.file(link, temp)
    df.data <- read.csv(unz(temp, "activity.csv"))
    unlink(temp)
} else
{
    df.data <- read.csv("./data/activity.csv")
}

dir.create(file.path('./figures'))
```

Two further columns are added to the data: a factor variable `day` that removes the year in date (for readibility reasons) and a factor variable `daytype` indicating if the day is a weekday or a weekend.  

```{r, echo=TRUE}
df.data <- mutate(df.data, 
                  day = factor(sub("2012-", "", df.data$date)),
                  daytype = factor(ifelse(weekdays(as.Date(df.data$date, format = "%Y-%m-%d")) %in% c("Saturday","Sunday"), "weekend", "weekday")))

knitr::kable(head(df.data))
```

## What is mean total number of steps taken per day

The following is the plot of the histogram of the total number of steps taken each day

```{r, echo=TRUE, message=FALSE}
df.data %>%
    group_by(date) %>%
    summarise(sum_step = sum(steps, na.rm = TRUE)) %>%
    ggplot(aes(x = sum_step)) +
        geom_histogram(color = "black", fill = "red")
    ggsave("./figures/sum_hist.png")
```

The mean and median of the total number of steps taken each day.

```{r, echo=TRUE, message=FALSE}
stepsum <- df.data %>% 
    group_by(date) %>%
    summarise(ssum = sum(steps, na.rm = TRUE))

mean.steps <- round(mean(stepsum$ssum),2)
median.steps <- median(stepsum$ssum)
```

- the mean is `mean.steps`: `r mean.steps` 
- the median is `median.steps`: `r median.steps`

# What is the average daily activity pattern


```{r, echo=TRUE, message=FALSE}
stepaverage <- df.data %>% 
    group_by(interval) %>%
    summarise(average = mean(steps, na.rm = TRUE))
    
ggplot(stepaverage, aes(interval, average)) +
    geom_line(color = "darkblue") +
    geom_point(shape = 'o')

ggsave("./figures/stepaverage.png")

misteps <- stepaverage[stepaverage$average == max(stepaverage$average),]$interval
```

- the five minute interval that, on average, contains the maximum number of steps is `misteps` = `r misteps` 


## Imputing missing values

The total number of missing values is calculated with:

```{r, echo=TRUE}
n.missing <- sum(is.na(df.data))
```

- `n.missing` = `r n.missing`

The strategy chosen for imputing missing variables is by considering the mean of the 5-mins intervals during the recorded 2 months and replace the missing value. `stepsumimp` is the new dataset, equal to the original used so far, that contains the new information.


```{r}
impute.mean <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))

stepsumimp <- df.data %>%
    group_by(interval) %>%
    mutate(
        imputedsteps = impute.mean(steps),
    )

knitr::kable(head(stepsumimp))
```

The histogram of the total number of steps taken each day after the missing values are imputed is the following:

```{r, message=FALSE}
stepsumimp %>%
    group_by(date) %>%
    summarise(sum_step_imp = sum(imputedsteps, na.rm = TRUE)) %>%
    ggplot(aes(x = sum_step_imp)) +
    geom_histogram(color = "black", fill = "red")

ggsave("./figures/sum_hist_imp.png")
```

## Are there differences in activity patterns between weekdays and weekends

A panel plot comparing the average number of steps per each interval across weekdays and weekends is produced with the following code:

```{r,message=FALSE}
dailypattern <- df.data %>%
    group_by(interval, daytype) %>%
    summarise(intervalav = mean(steps, na.rm = TRUE))

ggplot(dailypattern, aes(interval, intervalav)) +
    geom_line(color = "dodgerblue2") +
    facet_wrap(~daytype, ncol = 1)

ggsave("./figures/dailypattern.png")
```

The plot shows that on the weekends a slightly more actibity is performed in the afternoon (but maybe with a slower start).